---
title: "Data Requirements for rQSWATPlus"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Data Requirements for rQSWATPlus}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

## Overview

Setting up a SWAT+ model with `rQSWATPlus` requires several spatial datasets 
and lookup tables. This vignette describes each input, its required format, and 
best practices for data preparation.

## Required Raster Data

### Digital Elevation Model (DEM)

The DEM is the foundation for watershed delineation. It provides elevation 
values used to calculate flow directions, slopes, and drainage patterns.

**Format Requirements:**
- **File format**: GeoTIFF (`.tif`) recommended; any GDAL-supported raster format
- **Data type**: Numeric (integer or float)
- **CRS**: Projected coordinate system (e.g., UTM) strongly recommended
- **Units**: Elevation in meters preferred
- **Resolution**: Depends on watershed size; 30m (SRTM), 10m (NED), or 
  finer for small watersheds
- **Coverage**: Must cover the entire study watershed plus some buffer

```{r}
library(terra)
dem <- rast("path/to/dem.tif")
dem
#> class       : SpatRaster
#> dimensions  : 299, 209, 1  (nrow, ncol, nlyr)
#> resolution  : 30, 30  (x, y)
#> extent      : 567000, 573270, 6205000, 6213970  (xmin, xmax, ymin, ymax)
#> coord. ref. : ETRS89 / UTM zone 32N (EPSG:25832)
#> source      : dem.tif
#> name        : dem
```

**Sources:**
- [SRTM](https://earthexplorer.usgs.gov/) - 30m global
- [Copernicus DEM](https://spacedata.copernicus.eu/) - 30m global
- [USGS NED](https://apps.nationalmap.gov/) - 10m US
- National elevation services

**Tips:**
- Ensure the DEM covers the area beyond your watershed boundary (TauDEM needs 
  context for flow direction calculation)
- Fill any artifacts or voids before use (though TauDEM's pit removal handles 
  small depressions)
- Projected CRS is required for accurate area/slope calculations
- Use consistent units (meters) for both horizontal and vertical

### Land Use Raster

The land use raster classifies land cover types across your study area.

**Format Requirements:**
- **File format**: GeoTIFF recommended
- **Data type**: Integer (categorical values)
- **CRS**: Same as DEM (or will be reprojected)
- **Coverage**: Must cover the watershed
- **Values**: Integer codes that map to SWAT+ land use types via lookup table

```{r}
landuse <- rast("path/to/landuse.tif")
freq(landuse)
#>   layer value  count
#> 1     1     0  15234
#> 2     1  1100    567
#> 3     1  1120    123
#> ...
```

**Sources:**
- [CORINE Land Cover](https://land.copernicus.eu/) - Europe
- [NLCD](https://www.mrlc.gov/) - US
- [ESA WorldCover](https://esa-worldcover.org/) - 10m global
- National land use databases

### Soil Raster

The soil raster identifies soil types/map units across the study area.

**Format Requirements:**
- **File format**: GeoTIFF recommended
- **Data type**: Integer (soil map unit IDs)
- **CRS**: Same as DEM (or will be reprojected)
- **Values**: Integer codes mapping to soil names via lookup table

```{r}
soil <- rast("path/to/soil.tif")
freq(soil)
#>   layer value count
#> 1     1  1011  2345
#> 2     1  1018   678
#> ...
```

**Sources:**
- [STATSGO/SSURGO](https://websoilsurvey.nrcs.usda.gov/) - US
- [SoilGrids](https://soilgrids.org/) - Global
- [European Soil Database](https://esdac.jrc.ec.europa.eu/) - Europe
- National soil surveys

## Optional Vector Data

### Outlet Points (Shapefile)

Outlet points define where rivers exit the watershed. If provided, the 
watershed is delineated upstream of these points.

**Format Requirements:**
- **File format**: Shapefile (`.shp` with `.shx`, `.dbf`, `.prj`)
- **Geometry**: Point
- **CRS**: Same as DEM
- **Attributes**: Optional; IDs are auto-assigned

If no outlets are provided, TauDEM will delineate the entire DEM extent.

## Lookup Tables (CSV)

### Land Use Lookup

Maps integer raster values to SWAT+ land use codes.

**Format**: CSV with at least two columns

```
"Value","Landuse"
0000,PAST
1100,UTRN
1110,URHD
1120,URMD
1210,UIDU
2110,AGRL
...
```

The first column contains the raster integer value, the second contains the 
SWAT+ land use code.

**Common SWAT+ Land Use Codes:**

| Code | Description |
|------|-------------|
| AGRL | Agricultural land (generic) |
| AGRR | Agricultural row crops |
| CORN | Corn |
| SOYB | Soybeans |
| WHEA | Wheat |
| FRSD | Deciduous forest |
| FRSE | Evergreen forest |
| FRST | Mixed forest |
| PAST | Pasture |
| RNGE | Range/grassland |
| URBN | Urban (generic) |
| URHD | Urban high density |
| URMD | Urban medium density |
| UTRN | Urban transportation |
| WATR | Water |
| WETF | Forested wetland |
| WETL | Wetland |
| BARR | Barren |
| SWRN | Southwestern range |

### Soil Lookup

Maps integer raster values to soil names from the SWAT+ soils database.

**Format**: CSV with at least two columns

```
VALUE,NAME
1011,DK1011
1018,DK1018
1021,DK1021
...
```

Soil names should correspond to entries in your SWAT+ soil parameter database.

## Slope Class Definition

Slope classes subdivide HRUs by terrain steepness. They are defined using 
percentage breakpoints:

```{r}
library(rQSWATPlus)

# Single slope class (no subdivision by slope)
slopes <- qswat_create_slope_classes(c(0, 9999))

# Three slope classes: 0-5%, 5-15%, >15%
slopes <- qswat_create_slope_classes(c(0, 5, 15, 9999))

# Five slope classes
slopes <- qswat_create_slope_classes(c(0, 2, 5, 10, 25, 9999))
```

## Data Preparation Checklist

Before running `rQSWATPlus`, verify:

- [ ] **DEM**: GeoTIFF, projected CRS, covers watershed + buffer
- [ ] **Land use**: Integer raster, covers watershed
- [ ] **Soil**: Integer raster, covers watershed
- [ ] **Land use lookup**: CSV mapping all raster values to SWAT+ codes
- [ ] **Soil lookup**: CSV mapping all raster values to soil names
- [ ] **Outlets** (optional): Point shapefile in same CRS as DEM
- [ ] **CRS consistency**: All spatial data should share the same CRS 
  (reprojection is handled but may introduce artifacts)
- [ ] **Coverage**: All rasters cover the watershed extent
- [ ] **No data**: NoData values are properly set in all rasters

## Example: Inspecting the Included Ravn Dataset

```{r, eval = TRUE}
library(rQSWATPlus)

# Paths to included example data
dem_path <- system.file("extdata", "ravn_dem.tif", package = "rQSWATPlus")
lu_path <- system.file("extdata", "ravn_landuse.tif", package = "rQSWATPlus")
soil_path <- system.file("extdata", "ravn_soil.tif", package = "rQSWATPlus")
lu_csv <- system.file("extdata", "ravn_landuse.csv", package = "rQSWATPlus")
soil_csv <- system.file("extdata", "ravn_soil.csv", package = "rQSWATPlus")
```

```{r, eval = FALSE}
library(terra)

# Inspect DEM
dem <- rast(dem_path)
dem
plot(dem, main = "Ravn DEM")

# Check land use values and their lookup
lu <- rast(lu_path)
lu_lookup <- read.csv(lu_csv)
head(lu_lookup)

# Check soil values and their lookup
soil <- rast(soil_path)
soil_lookup <- read.csv(soil_csv)
head(soil_lookup)
```

## Coordinate Reference System Notes

- **Projected CRS** (UTM, State Plane, etc.) is required for accurate 
  calculations
- If your DEM is in geographic CRS (lat/lon), reproject it first:

```{r}
# Reproject DEM to UTM
dem_geo <- rast("dem_wgs84.tif")
dem_proj <- project(dem_geo, "EPSG:32632")  # UTM zone 32N
writeRaster(dem_proj, "dem_utm.tif")
```

- All rasters should ideally share the same CRS to avoid reprojection artifacts
- The package will attempt to align rasters internally, but pre-aligned data 
  gives best results
