---
title: "Introduction to rQSWATPlus"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to rQSWATPlus}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

## Overview

The `rQSWATPlus` package provides R functions to set up SWAT+ (Soil and Water 
Assessment Tool) hydrological models. It replicates the workflow of the 
[QSWATPlus](https://github.com/limnotrack/QSWATPlus) QGIS plugin, bringing 
SWAT+ model setup capabilities to R users.

SWAT+ is a watershed-scale model used for simulating hydrology, water quality, 
and land management impacts. Setting up a SWAT+ model requires several spatial 
processing steps that this package automates.

## Prerequisites

### TauDEM Installation

Watershed delineation in `rQSWATPlus` relies on 
[TauDEM](https://hydrology.usu.edu/taudem/taudem5/) (Terrain Analysis Using 
Digital Elevation Models) via the [`traudem`](https://github.com/lucarraro/traudem/) 
R package.

You need to install TauDEM and its dependencies (GDAL and MPI) on your system.
See the traudem installation vignette for platform-specific instructions:

```{r}
# Install traudem from CRAN
install.packages("traudem")

# Or the development version
# devtools::install_github("lucarraro/traudem")

# Check your TauDEM installation
traudem::taudem_sitrep()
```

### R Package Dependencies

```{r}
# Install rQSWATPlus
# devtools::install_github("limnotrack/QSWATPlus", subdir = "rQSWATPlus")

# Key dependencies (installed automatically):
# - terra: Raster data handling
# - sf: Vector/shapefile processing
# - RSQLite: Database output
# - traudem: TauDEM interface
```

## Quick Start

The basic workflow has five steps:

```{r}
library(rQSWATPlus)

# 1. Setup project
project <- qswat_setup(
  project_dir = "my_swatplus_project",
  dem_file = "path/to/dem.tif",
  landuse_file = "path/to/landuse.tif",
  soil_file = "path/to/soil.tif",
  landuse_lookup = "path/to/landuse_lookup.csv",
  soil_lookup = "path/to/soil_lookup.csv",
  outlet_file = "path/to/outlets.shp"  # optional
)

# 2. Delineate watershed
project <- qswat_delineate(project, threshold = 500)

# 3. Create stream network
project <- qswat_create_streams(project)

# 4. Create HRUs
lu_lookup <- qswat_read_landuse_lookup("path/to/landuse_lookup.csv")
soil_lookup <- qswat_read_soil_lookup("path/to/soil_lookup.csv")
slopes <- qswat_create_slope_classes(c(0, 5, 15, 9999))

project <- qswat_create_hrus(
  project, lu_lookup, soil_lookup, slopes
)

# 5. Write database
qswat_write_database(project)
```

Or use the convenience function for the entire workflow:

```{r}
project <- qswat_run(
  project_dir = "my_swatplus_project",
  dem_file = "path/to/dem.tif",
  landuse_file = "path/to/landuse.tif",
  soil_file = "path/to/soil.tif",
  landuse_lookup = "path/to/landuse_lookup.csv",
  soil_lookup = "path/to/soil_lookup.csv",
  threshold = 500,
  slope_breaks = c(0, 5, 15, 9999)
)
```

## Workflow Steps Explained

### 1. Project Setup

`qswat_setup()` creates a project directory structure and validates your input 
files. It checks that rasters are readable and copies source files into the 
project directory.

### 2. Watershed Delineation

`qswat_delineate()` runs the full TauDEM delineation workflow:

1. **Pit removal** (`pitremove`): Fill depressions in the DEM
2. **D8 flow direction** (`d8flowdir`): Calculate flow directions
3. **D-infinity flow direction** (`dinfflowdir`): Calculate slopes and angles
4. **Contributing area** (`aread8`): Compute flow accumulation
5. **Stream threshold** (`threshold`): Define stream/channel networks
6. **Move outlets** (`moveoutletstostreams`): Snap outlets to streams
7. **Stream network** (`streamnet`): Generate topology and watersheds

The `threshold` parameter controls the minimum drainage area (in cells) to 
define a stream. A lower threshold creates more streams and smaller subbasins.

### 3. Stream Network

`qswat_create_streams()` extracts the stream/channel topology from the TauDEM 
output, including stream links, downstream connections, stream orders, and 
drainage areas.

### 4. HRU Creation

`qswat_create_hrus()` overlays land use, soil, and slope data on the delineated 
watershed. Each unique combination of subbasin × land use × soil × slope class 
creates a Hydrologic Response Unit (HRU).

Area thresholds can be used to filter out small HRUs:
- `landuse_threshold`: Minimum % of subbasin area for a land use
- `soil_threshold`: Minimum % of land use area for a soil type
- `slope_threshold`: Minimum % of soil area for a slope class

### 5. Database Output

`qswat_write_database()` creates a SQLite database compatible with the SWAT+ 
Editor for further parameterization and model execution.

## Next Steps

- See `vignette("data-requirements")` for detailed input data format 
  specifications
- See `vignette("workflow")` for a complete worked example using the included 
  Ravn watershed dataset
