---
title: "Complete Workflow Example"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Complete Workflow Example}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = TRUE
)
```

## Overview

This vignette demonstrates a complete SWAT+ model setup using `rQSWATPlus` 
with the included Ravn watershed dataset from Denmark. The workflow covers 
every step from project initialization to database creation.

## The Ravn Watershed Dataset

The Ravn watershed is a small catchment in Denmark included as example data.
It includes:

- **DEM**: 30m resolution SRTM elevation data
- **Land use**: Categorical raster with 55+ land cover classes
- **Soil**: Categorical raster with 49 soil types
- **Outlet**: Point shapefile marking the watershed outlet
- **Lookup tables**: CSV files mapping raster values to SWAT+ codes

## Step 1: Load Package and Check Prerequisites

```{r setup}
library(rQSWATPlus)

# Verify TauDEM is available
traudem::taudem_sitrep()
```

## Step 2: Locate Example Data

```{r data-paths}
# Get paths to example data included in the package
dem <- system.file("extdata", "ravn_dem.tif", package = "rQSWATPlus")
landuse <- system.file("extdata", "ravn_landuse.tif", package = "rQSWATPlus")
soil <- system.file("extdata", "ravn_soil.tif", package = "rQSWATPlus")
lu_lookup <- system.file("extdata", "ravn_landuse.csv", package = "rQSWATPlus")
soil_lookup <- system.file("extdata", "ravn_soil.csv", package = "rQSWATPlus")
outlet <- system.file("extdata", "ravn_outlet.shp", package = "rQSWATPlus")
```

## Step 3: Inspect the Input Data

It is good practice to inspect your input data before running the workflow.

```{r inspect}
library(terra)

# Inspect the DEM
dem_rast <- rast(dem)
dem_rast

# Check dimensions and resolution
cat("DEM dimensions:", nrow(dem_rast), "x", ncol(dem_rast), "\n")
cat("Resolution:", res(dem_rast), "\n")
cat("CRS:", crs(dem_rast, describe = TRUE)$name, "\n")

# Inspect land use raster
lu_rast <- rast(landuse)
cat("\nLand use unique values:", length(unique(values(lu_rast, na.rm = TRUE))), "\n")

# Read and inspect lookup tables
lu_table <- read.csv(lu_lookup)
cat("\nLand use lookup (first 10 rows):\n")
print(head(lu_table, 10))

soil_table <- read.csv(soil_lookup)
cat("\nSoil lookup (first 10 rows):\n")
print(head(soil_table, 10))
```

## Step 4: Set Up the Project

```{r setup-project}
# Create a project directory (use a temporary directory for this example)
project_dir <- file.path(tempdir(), "ravn_swatplus")

# Initialize the project
project <- qswat_setup(
  project_dir = project_dir,
  dem_file = dem,
  landuse_file = landuse,
  soil_file = soil,
  landuse_lookup = lu_lookup,
  soil_lookup = soil_lookup,
  outlet_file = outlet
)

# Print project summary
print(project)
```

## Step 5: Run Watershed Delineation

The delineation step processes the DEM using TauDEM to identify streams, 
channels, subbasins, and watershed boundaries.

The `threshold` parameter controls stream definition - it specifies the minimum 
number of cells that must drain to a point for it to be classified as a stream. 
A lower threshold creates more streams and smaller subbasins.

```{r delineate}
# Run delineation with a stream threshold of 500 cells
# This means a cell needs at least 500 upstream cells to be a stream
project <- qswat_delineate(
  project,
  threshold = 500,         # Stream threshold (cells)
  channel_threshold = 50,  # Channel threshold (finer network)
  quiet = TRUE            # Show TauDEM progress messages
)
```

### Understanding the Delineation Steps

The delineation runs these TauDEM operations in sequence:

1. **Pit Removal** (`pitremove`): Fills depressions in the DEM that would 
   trap water and prevent proper flow routing
   
2. **D8 Flow Direction** (`d8flowdir`): Assigns each cell one of 8 flow 
   directions pointing to the steepest downslope neighbor
   
3. **D-infinity Flow Direction** (`dinfflowdir`): Calculates continuous flow 
   angles and slopes (used for landscape analysis)
   
4. **Contributing Area** (`aread8`): Computes the number of upstream cells 
   draining through each cell
   
5. **Threshold** (`threshold`): Marks cells as streams where contributing 
   area exceeds the threshold value
   
6. **Move Outlets** (`moveoutletstostreams`): Snaps user-provided outlet 
   points to the nearest stream cell
   
7. **Stream Network** (`streamnet`): Generates the complete stream network 
   topology, watershed boundaries, and subbasin delineation

## Step 6: Create Stream Network

Extract stream topology from the delineation results.

```{r streams}
project <- qswat_create_streams(project)

# View stream topology
head(project$stream_topology)
```

## Step 7: Create HRUs

HRUs (Hydrologic Response Units) are the fundamental computational units in 
SWAT+. Each HRU is a unique combination of subbasin, land use, soil type, 
and slope class.

```{r hrus}
# Read lookup tables
lu <- qswat_read_landuse_lookup(lu_lookup)
soil_lkp <- qswat_read_soil_lookup(soil_lookup)

# Define slope classes: 0-5%, 5-15%, and >15%
slopes <- qswat_create_slope_classes(c(0, 5, 15, 9999))
print(slopes)

# Create HRUs with area thresholds to filter small units
project <- qswat_create_hrus(
  project,
  landuse_lookup = lu,
  soil_lookup = soil_lkp,
  slope_classes = slopes,
  landuse_threshold = 5,   # Land use must be >5% of subbasin
  soil_threshold = 5,      # Soil must be >5% of land use area
  slope_threshold = 0       # No slope filtering
)

# View HRU summary
cat("Total HRUs:", nrow(project$hru_data), "\n")
cat("Subbasins:", length(unique(project$hru_data$subbasin)), "\n")

# View first few HRUs
head(project$hru_data)

# View basin statistics
print(project$basin_data)
```

### Understanding HRU Thresholds

The threshold filtering uses a "dominant" approach:

1. **Land use threshold** (e.g., 5%): Within each subbasin, only land uses 
   covering more than 5% of the subbasin area are kept.

2. **Soil threshold** (e.g., 5%): Within each remaining land use, only soils 
   covering more than 5% of that land use's area are kept.

3. **Slope threshold** (e.g., 10%): Within each remaining soil type, only 
   slope classes covering more than 10% of that soil's area are kept.

Higher thresholds reduce the number of HRUs but may oversimplify the landscape.

## Step 8: Write the SWAT+ Database

```{r database}
# Write the project database
db_path <- qswat_write_database(project, overwrite = TRUE)

# Verify the database
library(DBI)
library(RSQLite)
con <- dbConnect(SQLite(), db_path)

# List tables
cat("Database tables:\n")
print(dbListTables(con))

# Check table contents
cat("\nSubbasins:\n")
print(dbGetQuery(con, "SELECT * FROM gis_subbasins"))

cat("\nHRUs (first 10):\n")
print(dbGetQuery(con, "SELECT * FROM gis_hrus LIMIT 10"))

cat("\nRouting:\n")
print(dbGetQuery(con, "SELECT * FROM gis_routing"))

dbDisconnect(con)
```

## Alternative: Run Everything at Once

For convenience, `qswat_run()` executes the entire workflow:

```{r run-all}
project <- qswat_run(
  project_dir = file.path(tempdir(), "ravn_quick"),
  dem_file = dem,
  landuse_file = landuse,
  soil_file = soil,
  landuse_lookup = lu_lookup,
  soil_lookup = soil_lookup,
  outlet_file = outlet,
  threshold = 500,
  slope_breaks = c(0, 5, 15, 9999),
  landuse_threshold = 5,
  soil_threshold = 5
)
```

## Next Steps After rQSWATPlus

After creating the project database with `rQSWATPlus`, the next steps in the 
SWAT+ modeling workflow are:

1. **SWAT+ Editor**: Open the SQLite database in the SWAT+ Editor to:
   - Add weather/climate data
   - Configure management operations
   - Set up point sources
   - Configure reservoirs and ponds
   - Write SWAT+ input files

2. **SWAT+ Simulation**: Run the SWAT+ executable with the generated input files

3. **Output Analysis**: Analyze SWAT+ outputs for hydrology, water quality, 
   sediment transport, etc.

## Cleanup

```{r cleanup}
# Remove temporary project directories
unlink(file.path(tempdir(), "ravn_swatplus"), recursive = TRUE)
unlink(file.path(tempdir(), "ravn_quick"), recursive = TRUE)
```
